---
title: "eSVD2: `master` (3d5f7bf) versus `devel` (1.1.0) on simulated cohorts"
author: "Drafted by Claude for Kevin Z. Lin"
date: "2026-09-29"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    theme: flatly
    highlight: tango
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.align = "center")
```

```{r libraries}
library(ggplot2)
```

```{r read_data}
out_dir <- "output"
version_vec <- c("master", "devel", "devel_swap")

gene_df <- do.call(rbind, lapply(version_vec, function(v){
  utils::read.csv(file.path(out_dir, v, "genes.csv"))
}))
truth_df <- utils::read.csv(file.path(out_dir, "truth.csv"))
gene_df <- merge(gene_df, truth_df, by = c("regime", "rep", "gene"))
gene_df$p <- 10^(-gene_df$log10p)
gene_df$status <- factor(ifelse(gene_df$is_de, "DE", "null"),
                         levels = c("null", "DE"))

run_df <- do.call(rbind, lapply(version_vec, function(v){
  utils::read.csv(file.path(out_dir, v, "runs.csv"))
}))
corner_df <- do.call(rbind, lapply(c("master", "devel"), function(v){
  utils::read.csv(file.path(out_dir, v, "corner_cases.csv"))
}))
wall_time_df <- utils::read.csv(file.path(out_dir, "wall_time.csv"),
                                header = FALSE)

# One row per gene per data set, with the versions side by side.
.wide <- function(column){
  master_df <- gene_df[gene_df$version == "master",
                       c("regime", "rep", "gene", "status", column)]
  devel_df <- gene_df[gene_df$version == "devel",
                      c("regime", "rep", "gene", column)]
  swap_df <- gene_df[gene_df$version == "devel_swap",
                     c("regime", "rep", "gene", column)]
  colnames(master_df)[5] <- "master"
  colnames(devel_df)[4] <- "devel"
  colnames(swap_df)[4] <- "devel_swap"
  res <- merge(master_df, devel_df, by = c("regime", "rep", "gene"))
  merge(res, swap_df, by = c("regime", "rep", "gene"))
}

regime_levels <- c("baseline", "null", "strong_de", "near_poisson",
                   "minimal_design", "generate_null")
version_col_vec <- c(master = "#2a78d6", devel = "#eb6834",
                     devel_swap = "#1baf7a")
status_col_vec <- c(null = "#9e9d97", DE = "#4a3aa7")

theme_set(theme_bw(base_size = 11))
theme_update(panel.grid.minor = element_blank(),
             strip.background = element_rect(fill = "#f1f0ec", colour = NA))
```

```{r headline_numbers}
num_null_genes <- sum(!gene_df$is_de & gene_df$version == "master" &
                        gene_df$regime == "null") / 10
.fp_per_rep <- function(v, r){
  idx <- gene_df$version == v & gene_df$regime == r
  sum(gene_df$fdr[idx] < 0.05 & !gene_df$is_de[idx]) / 10
}
.type1 <- function(v, r){
  idx <- gene_df$version == v & gene_df$regime == r & !gene_df$is_de
  mean(gene_df$p[idx] < 0.05)
}
logfc_wide_df <- .wide("logFC")
min_logfc_cor <- min(sapply(split(logfc_wide_df, logfc_wide_df$regime),
                            function(x){stats::cor(x$master, x$devel)}))
teststat_wide_df <- .wide("teststat")
swap_cor <- stats::cor(teststat_wide_df$master, teststat_wide_df$devel_swap)
```

# Overview

This report compares two versions of `eSVD2` on the same simulated data:

* **`master`**: git commit `3d5f7bf`, `DESCRIPTION` version 1.0.1.07. This is
  what `devtools::install_github("linnykos/eSVD2")` installs today.
* **`devel`**: the working tree on the `devel` branch (commit `d49e402`),
  version 1.1.0, the candidate for CRAN.

It asks two questions. **Which behaviors got better?** And **did the rewrite
change what eSVD-DE reports on ordinary data, or mostly how it behaves at the
edges?**

**The short answer.**

1. **The point estimate did not change.** The log2 fold changes of the two
   versions agree at a Pearson correlation of at least
   `r sprintf("%.4f", min_logfc_cor)` in every regime.
2. **Most of the fixes are corner-case fixes, and they behave as `NEWS.md`
   says.** Inputs that used to crash deep in the optimizer with
   `missing value where TRUE/FALSE needed`, or finish silently with a wrong
   answer, now stop at the first step with a message that names the
   problem (section 7).
3. **One change does move ordinary results: the overdispersion estimator
   (`gamma_rate`).** It changes every gene's nuisance rate, so it changes
   every Welch statistic and p-value. Replacing `devel`'s nuisance estimates
   with `master`'s reproduces `master`'s Welch statistics at correlation
   `r sprintf("%.4f", swap_cor)` (section 5.3). In these simulations the new
   estimator ranks genes by their true overdispersion far better than the old
   one, but it produces **more false discoveries**. Under the global null,
   `r .fp_per_rep("devel", "null")` of `r num_null_genes` null genes per data set
   reach FDR < 0.05, against `r .fp_per_rep("master", "null")` under `master`.
   In the near-Poisson regime the type-I error at p < 0.05 is
   `r sprintf("%.2f", .type1("devel", "near_poisson"))` against
   `r sprintf("%.2f", .type1("master", "near_poisson"))`. **This is a
   dramatic change, and it is not an improvement** (section 6).

## How the comparison was run

The two versions are installed into separate private libraries and run in
separate R processes. The step-by-step API (`initialize_esvd()` through
`compute_pvalue()`) has the same function names and arguments in both, so
one script drives either version, calling the steps in the order `eSVD()`
runs them. All heavy computation happens in the scripts; this document only
reads their CSVs.

| Script | What it does |
|---|---|
| `00_install_versions_claude.sh` | `git archive 3d5f7bf` → `lib/master`; the working tree → `lib/devel` |
| `helpers_claude.R` | the simulator and the pipeline, shared by the scripts below |
| `01_simulate_data_claude.R` | 6 regimes × 10 replicates → `data/*.rds`, and the truth → `output/truth.csv` |
| `02_run_regimes_claude.R` | fits all 60 data sets with one version → `output/<version>/genes.csv`, `runs.csv` |
| `03_corner_cases_claude.R` | 12 perturbed cohorts per version → `output/<version>/corner_cases.csv` |
| `run_all_claude.sh` | all of the above, end to end |

The whole of `run_all_claude.sh` took **`r round(wall_time_df[1, 2] / 60, 1)`
minutes** on the laptop it was run on.

**`devel_swap` is an ablation, not a third version.** It runs the `devel` code
on each data set, but right after `estimate_nuisance()` it overwrites the
per-gene nuisance rate with the value `master` estimated on the same data set.
Where `devel_swap` matches `master`, the nuisance estimate explains the
difference between the versions.

## The simulated regimes

Every regime has 20 individuals (10 case, 10 control), 30 cells each, and 300
genes. Except for `generate_null`, the data come from the eSVD hierarchical
model itself, using the generator of `.build_tiny_data()` in
`tests/testthat/helper-fixtures.R`. That model has a rank-3 latent structure,
an individual-level random effect, Sex and Age effects, a per-cell library
size, and counts $A_{ji} \sim \mathrm{Poisson}(\ell_{ji} \lambda_{ji})$ with
$\lambda_{ji} \sim \mathrm{Gamma}(\text{mean} = \mu_{ji}, \text{rate} = \beta_j)$.
The truth for a gene's log2 fold change is the ratio of the arithmetic
means of $\exp(\text{natural parameter})$ in the two arms, which is what
both the logFC and the Welch statistic contrast.

| Regime | DE genes | Nuisance rate $\beta_j$ | What it probes |
|---|---|---|---|
| `baseline` | 15, effect 0.8 (natural log) | U(2, 8) on the log scale | ordinary data |
| `null` | 0 | U(2, 8) | calibration |
| `strong_de` | 15, effect ±3 | U(2, 8) | large Welch statistics; the t → z mapping |
| `near_poisson` | 15, effect 0.8 | U(50, 1000) | little overdispersion; the old `gamma_rate` cap |
| `minimal_design` | 15, effect 0.8 | U(2, 8) | no Sex/Age, so covariates are only Intercept + Log_UMI + CC (the GLM-fallback offset fix) |
| `generate_null` | 10 | the package's own `generate_null()` | a generator not written for this report |

Each regime has 10 replicates with different seeds. The fit uses `k = 3`,
`l2pen = 0.1`, `max_iter = 50`, and otherwise the package defaults.

# What changed, according to `NEWS.md`

| Change in `devel` | Where this report looks at it |
|---|---|
| Welch → Gaussian mapping done on the log scale, so a large statistic no longer becomes `Inf` | 6.4 |
| `multtest()` reports its empirical-null estimator and warns on the `locfdr` fallback; Efron's truncated MLE regains `p0 <= 1` | 6.5, 7 |
| `gamma_rate` no longer caps its search at the largest library size | 4, 5, 6 |
| `initialize_esvd()` refuses all-zero genes; zeroes `NA` in sparse input | 7 |
| GLM fallback (≤ 1 covariate besides intercept and offsets) keeps the library-size offset | 3, regime `minimal_design` |
| Collinear covariates refused by name | 7 |
| `eSVD()` refuses individuals in both arms, one-individual arms, and individuals with too few cells | 7 |
| Iterative SVD starts from a fixed vector (determinism) | 7 |
| New: `log2fc_vec`, `log2fc_se_vec`, `compute_log_fold_change()` | 8 |

# Point estimates: the log2 fold change

```{r logfc_scatter, fig.width=9, fig.height=5.5}
plot_df <- logfc_wide_df
plot_df$regime <- factor(plot_df$regime, levels = regime_levels)
plot1 <- ggplot2::ggplot(plot_df, ggplot2::aes(x = master, y = devel))
plot1 <- plot1 + ggplot2::geom_abline(slope = 1, intercept = 0,
                                      colour = "#c3c2b7", linewidth = 0.5)
plot1 <- plot1 + ggplot2::geom_point(ggplot2::aes(colour = status),
                                     size = 0.8, alpha = 0.6)
plot1 <- plot1 + ggplot2::facet_wrap(~ regime, nrow = 2, scales = "free")
plot1 <- plot1 + ggplot2::scale_colour_manual(values = status_col_vec,
                                              name = "True status")
plot1 <- plot1 + ggplot2::labs(x = "master: log2 fold change",
                               y = "devel: log2 fold change",
                               title = "The log2 fold change is essentially unchanged",
                               subtitle = "All 10 replicates per regime; grey line is y = x")
plot(plot1)
```

```{r logfc_table}
agreement_df <- do.call(rbind, lapply(regime_levels, function(r){
  x <- logfc_wide_df[logfc_wide_df$regime == r, ]
  data.frame(regime = r,
             pearson = stats::cor(x$master, x$devel),
             median_abs_diff = stats::median(abs(x$master - x$devel)),
             max_abs_diff = max(abs(x$master - x$devel)))
}))
knitr::kable(agreement_df, digits = 4,
             col.names = c("Regime", "Pearson r", "Median |Δ|", "Max |Δ|"),
             caption = "log2 fold change, master vs devel, all genes and replicates.")
```

The logFC is `log2(case_mean / control_mean)`, where each mean averages the
per-individual posterior means. The posterior mean moves only a little with
the nuisance rate, so this estimate is nearly identical across versions. The
small differences come from the fit and from the rescaling by
`bool_stabilize_underdispersion`. `devel` returns this quantity directly as
`log2fc_vec`; for `master` it is computed from `case_mean` and `control_mean`.
The two definitions are the same.

**`minimal_design` is where the GLM-fallback offset fix would show, and it
does not.** With only Intercept + Log_UMI + CC, `master` fitted the intercepts
without the depth offset. That changes the initialization, but after two
rounds of `opt_esvd()` the fold changes agree as closely as in any other
regime.

# The nuisance (overdispersion) estimate

`nuisance_vec` is the Gamma **rate** $\beta_j = 1/\gamma_j$: a large value
means little overdispersion, and $\beta_j \to \infty$ is a Poisson gene.

## Against the truth

```{r nuisance_truth, fig.width=9, fig.height=6}
plot_df <- gene_df[gene_df$version %in% c("master", "devel") &
                     gene_df$regime != "generate_null", ]
plot_df$regime <- factor(plot_df$regime, levels = regime_levels)
plot_df$version <- factor(plot_df$version, levels = c("master", "devel"))
plot1 <- ggplot2::ggplot(plot_df, ggplot2::aes(x = nuisance_true, y = nuisance))
plot1 <- plot1 + ggplot2::geom_abline(slope = 1, intercept = 0,
                                      colour = "#52514e", linewidth = 0.5,
                                      linetype = "dashed")
plot1 <- plot1 + ggplot2::geom_point(ggplot2::aes(colour = version),
                                     size = 0.7, alpha = 0.45)
plot1 <- plot1 + ggplot2::facet_grid(version ~ regime, scales = "free_x")
plot1 <- plot1 + ggplot2::scale_x_log10()
plot1 <- plot1 + ggplot2::scale_y_log10()
plot1 <- plot1 + ggplot2::scale_colour_manual(values = version_col_vec,
                                              guide = "none")
plot1 <- plot1 + ggplot2::labs(x = "True rate (log scale)",
                               y = "Estimated rate (log scale)",
                               title = "master's estimate carries no information about the truth; devel's ranks it but diverges",
                               subtitle = "Dashed line is y = x. Points near 1e7 are devel estimates that ran off to the upper end of the search.")
plot(plot1)
```

```{r nuisance_table}
nuisance_acc_df <- do.call(rbind, lapply(setdiff(regime_levels, "generate_null"),
                                         function(r){
  do.call(rbind, lapply(c("master", "devel"), function(v){
    x <- gene_df[gene_df$regime == r & gene_df$version == v, ]
    data.frame(regime = r,
               version = v,
               spearman = stats::cor(x$nuisance, x$nuisance_true,
                                     method = "spearman"),
               median_ratio = stats::median(x$nuisance / x$nuisance_true),
               max_estimate = max(x$nuisance),
               frac_diverged = mean(x$nuisance > 1e4))
  }))
}))
knitr::kable(nuisance_acc_df, digits = c(0, 0, 2, 2, 0, 3),
             col.names = c("Regime", "Version", "Spearman with truth",
                           "Median estimate / truth", "Largest estimate",
                           "Fraction > 1e4"),
             caption = "Nuisance rate against the truth, all genes and replicates.")
```

What the plot and table show:

* **`master`'s estimates are nearly independent of the truth** (Spearman
  near 0) and never exceed a few tens: the old search interval ended at
  the largest library size. Even in `near_poisson`, where the true rates are
  50 to 1000, `master` reports rates in the single digits and tens. It
  therefore treats every gene as moderately overdispersed.
* **`devel` ranks the genes correctly** (Spearman about 0.6 when the truth is
  U(2, 8)), **but it is biased upward two- to threefold, and some estimates
  diverge.** About 0.5% of ordinary genes, and about half of the
  `near_poisson` genes, land near $10^7$. This is Q10 / Q-LFC-1 in
  `CLAUDE_kevin.md`, now with a measured consequence.

## Why the nuisance matters downstream

The posterior shrinks each cell's observed count toward the fitted mean with
weight governed by $\beta_j$. A huge $\beta_j$ makes the posterior collapse onto
the fitted mean. The within-individual posterior variance then vanishes, the
denominator of the Welch statistic shrinks, and the statistic grows. The next
section follows that chain through to the p-values.

# Test statistics and p-values

## Welch statistic, master vs devel

```{r teststat_scatter, fig.width=9, fig.height=5.5}
plot_df <- teststat_wide_df
plot_df$regime <- factor(plot_df$regime, levels = regime_levels)
plot1 <- ggplot2::ggplot(plot_df, ggplot2::aes(x = master, y = devel))
plot1 <- plot1 + ggplot2::geom_abline(slope = 1, intercept = 0,
                                      colour = "#c3c2b7", linewidth = 0.5)
plot1 <- plot1 + ggplot2::geom_point(ggplot2::aes(colour = status),
                                     size = 0.8, alpha = 0.6)
plot1 <- plot1 + ggplot2::facet_wrap(~ regime, nrow = 2, scales = "free")
plot1 <- plot1 + ggplot2::scale_colour_manual(values = status_col_vec,
                                              name = "True status")
plot1 <- plot1 + ggplot2::labs(x = "master: Welch statistic",
                               y = "devel: Welch statistic",
                               title = "Most genes sit on y = x; a few leave it by an order of magnitude",
                               subtitle = "Grey line is y = x. The genes far off the line are those whose devel nuisance rate diverged (section 6.3).")
plot(plot1)
```

## −log10 p-value, master vs devel

```{r pvalue_scatter, fig.width=9, fig.height=5.5}
plot_df <- .wide("log10p")
plot_df$regime <- factor(plot_df$regime, levels = regime_levels)
plot1 <- ggplot2::ggplot(plot_df, ggplot2::aes(x = master, y = devel))
plot1 <- plot1 + ggplot2::geom_abline(slope = 1, intercept = 0,
                                      colour = "#c3c2b7", linewidth = 0.5)
plot1 <- plot1 + ggplot2::geom_point(ggplot2::aes(colour = status),
                                     size = 0.8, alpha = 0.6)
plot1 <- plot1 + ggplot2::facet_wrap(~ regime, nrow = 2, scales = "free")
plot1 <- plot1 + ggplot2::scale_colour_manual(values = status_col_vec,
                                              name = "True status")
plot1 <- plot1 + ggplot2::labs(x = "master: −log10 p", y = "devel: −log10 p")
plot(plot1)
```

```{r teststat_table}
stat_agree_df <- do.call(rbind, lapply(regime_levels, function(r){
  x <- teststat_wide_df[teststat_wide_df$regime == r, ]
  y <- .wide("log10p")
  y <- y[y$regime == r, ]
  data.frame(regime = r,
             teststat_cor = stats::cor(x$master, x$devel),
             teststat_median = stats::median(abs(x$master - x$devel)),
             swap_cor = stats::cor(x$master, x$devel_swap),
             swap_max = max(abs(x$master - x$devel_swap)),
             log10p_spearman = stats::cor(y$master, y$devel,
                                          method = "spearman"))
}))
knitr::kable(stat_agree_df, digits = 4,
             col.names = c("Regime", "Welch r (master, devel)",
                           "Median |Δ| (master, devel)",
                           "Welch r (master, devel_swap)",
                           "Max |Δ| (master, devel_swap)",
                           "−log10 p Spearman (master, devel)"),
             caption = "Agreement of test statistics; all genes and replicates.")
```

The Pearson correlations in the second column understate the agreement for
most genes. The typical gene moves little (median |Δ| in the third column),
and the correlation is pulled down by the few genes that move a great deal.

## The ablation: the nuisance estimate is the whole difference

The table's third column is the decisive one. **Put `master`'s nuisance rates
into the `devel` code and the Welch statistics match `master`'s in every regime**: to within 0.006, and to all six stored
digits in four of the six regimes. The other changes between the versions (the
reparameterization by QR, the offset in the GLM fallback, the deterministic
SVD start, the rewritten empirical-null estimators) move nothing that reaches
the statistic in these regimes. Everything that differs between the versions'
Welch statistics comes from the one line that changed `gamma_rate`'s search
interval.

```{r swap_scatter, fig.width=9, fig.height=3.5}
plot_df <- teststat_wide_df[teststat_wide_df$regime %in%
                              c("baseline", "null", "near_poisson"), ]
plot_df$regime <- factor(plot_df$regime, levels = regime_levels)
plot1 <- ggplot2::ggplot(plot_df, ggplot2::aes(x = master, y = devel_swap))
plot1 <- plot1 + ggplot2::geom_abline(slope = 1, intercept = 0,
                                      colour = "#c3c2b7", linewidth = 0.5)
plot1 <- plot1 + ggplot2::geom_point(colour = version_col_vec[["devel_swap"]],
                                     size = 0.8, alpha = 0.6)
plot1 <- plot1 + ggplot2::facet_wrap(~ regime, nrow = 1, scales = "free")
plot1 <- plot1 + ggplot2::labs(x = "master: Welch statistic",
                               y = "devel with master's nuisance: Welch statistic",
                               title = "devel_swap reproduces master exactly")
plot(plot1)
```

# Calibration and power

## Null genes: p-value QQ plots

The p-values below are the ones each version reports, measured against its
estimated empirical null. Each panel pools the null genes of all 10 replicates.

```{r qq_plot, fig.width=9, fig.height=5.5}
qq_list <- list()
for(r in regime_levels){
  for(v in c("master", "devel")){
    p_vec <- gene_df$p[gene_df$regime == r & gene_df$version == v &
                         !gene_df$is_de]
    p_vec <- sort(p_vec)
    qq_list[[paste(r, v)]] <- data.frame(regime = r,
                                         version = v,
                                         expected = -log10(stats::ppoints(length(p_vec))),
                                         observed = -log10(p_vec))
  }
}
qq_df <- do.call(rbind, qq_list)
qq_df$regime <- factor(qq_df$regime, levels = regime_levels)
qq_df$version <- factor(qq_df$version, levels = c("master", "devel"))
plot1 <- ggplot2::ggplot(qq_df, ggplot2::aes(x = expected, y = observed))
plot1 <- plot1 + ggplot2::geom_abline(slope = 1, intercept = 0,
                                      colour = "#52514e", linewidth = 0.5,
                                      linetype = "dashed")
plot1 <- plot1 + ggplot2::geom_point(ggplot2::aes(colour = version),
                                     size = 0.8, alpha = 0.7)
plot1 <- plot1 + ggplot2::facet_wrap(~ regime, nrow = 2, scales = "free_y")
plot1 <- plot1 + ggplot2::scale_colour_manual(values = version_col_vec,
                                              name = "Version")
plot1 <- plot1 + ggplot2::labs(x = "Expected −log10 p (uniform)",
                               y = "Observed −log10 p, null genes",
                               title = "devel's null p-values have a heavier upper tail; near_poisson is badly miscalibrated")
plot(plot1)
```

## False and true discoveries at FDR < 0.05

```{r discovery_table}
discovery_df <- do.call(rbind, lapply(regime_levels, function(r){
  do.call(rbind, lapply(version_vec, function(v){
    x <- gene_df[gene_df$regime == r & gene_df$version == v, ]
    sig_vec <- x$fdr < 0.05
    data.frame(regime = r,
               version = v,
               type1 = mean(x$p[!x$is_de] < 0.05),
               fp = sum(sig_vec & !x$is_de) / 10,
               tp = sum(sig_vec & x$is_de) / 10,
               num_de = sum(x$is_de) / 10,
               fdp = sum(sig_vec & !x$is_de) / max(1, sum(sig_vec)))
  }))
}))
knitr::kable(discovery_df, digits = c(0, 0, 3, 1, 1, 0, 3),
             col.names = c("Regime", "Version", "Type-I at p < 0.05",
                           "False discoveries / data set",
                           "True discoveries / data set", "DE genes",
                           "Pooled FDP"),
             caption = "Discoveries at FDR < 0.05 (BH on the empirical-null p-values), averaged over 10 replicates of 300 genes.")
```

```{r fp_plot, fig.width=9, fig.height=3.8}
fp_list <- list()
for(r in regime_levels){
  for(v in c("master", "devel")){
    for(i in 1:10){
      idx <- gene_df$regime == r & gene_df$version == v & gene_df$rep == i
      fp_list[[paste(r, v, i)]] <- data.frame(regime = r,
                                              version = v,
                                              rep = i,
                                              fp = sum(gene_df$fdr[idx] < 0.05 &
                                                         !gene_df$is_de[idx]))
    }
  }
}
fp_df <- do.call(rbind, fp_list)
fp_df$regime <- factor(fp_df$regime, levels = regime_levels)
fp_df$version <- factor(fp_df$version, levels = c("master", "devel"))
plot1 <- ggplot2::ggplot(fp_df, ggplot2::aes(x = version, y = fp + 1))
plot1 <- plot1 + ggplot2::geom_point(ggplot2::aes(colour = version),
                                     position = ggplot2::position_jitter(width = 0.15,
                                                                         height = 0,
                                                                         seed = 1),
                                     size = 2)
plot1 <- plot1 + ggplot2::facet_wrap(~ regime, nrow = 1)
plot1 <- plot1 + ggplot2::scale_y_log10()
plot1 <- plot1 + ggplot2::scale_colour_manual(values = version_col_vec,
                                              guide = "none")
plot1 <- plot1 + ggplot2::labs(x = NULL,
                               y = "False discoveries + 1 (log scale)",
                               title = "False discoveries per data set at FDR < 0.05",
                               subtitle = "One point per replicate")
plot(plot1)
```

What this shows:

* **At p < 0.05 the null genes are calibrated in both versions in the ordinary
  regimes** (type-I error 0.04 to 0.06). The empirical null absorbs most of
  the shift in the statistics.
* **The difference is in the far tail, which is where FDR control acts.**
  `devel`'s null statistics have a heavier upper tail, so a few null genes
  per data set pass BH. About a quarter to a half of those (all of them in
  `near_poisson`) are genes whose
  nuisance rate diverged (section 6.3). The rest reflect the general upward
  bias of `devel`'s rates.
* **`near_poisson` is badly miscalibrated under `devel`**: about 60 of 285
  null genes per data set are declared DE, and power also drops. In that
  regime about half the genes' rates diverge, the empirical null is fitted to
  a mixture of collapsed and uncollapsed genes, and neither group is scored
  correctly.
* **`master` is not "right" here either.** Its rates ignore the truth, but
  treating every gene as moderately overdispersed happens to be
  conservative, so its errors are errors of caution.

## Where devel's false discoveries come from

**A diverged rate is close to a guaranteed false discovery.** In the ordinary
regimes, 74% to 100% of the null genes whose rate diverged reach FDR < 0.05,
against under 1% of the others (table below). The divergence is rare
(about 0.5% of genes), but each occurrence is almost always an error.

```{r near_poisson_mechanism, fig.width=9, fig.height=4}
plot_df <- gene_df[gene_df$version == "devel" &
                     gene_df$regime %in% c("null", "near_poisson"), ]
plot_df$regime <- factor(plot_df$regime, levels = regime_levels)
plot_df$discovery <- ifelse(plot_df$fdr < 0.05, "FDR < 0.05", "not significant")
plot1 <- ggplot2::ggplot(plot_df, ggplot2::aes(x = nuisance, y = abs(teststat)))
plot1 <- plot1 + ggplot2::geom_point(ggplot2::aes(colour = discovery,
                                                  shape = status),
                                     size = 1.2, alpha = 0.7)
plot1 <- plot1 + ggplot2::facet_wrap(~ regime, nrow = 1, scales = "free_y")
plot1 <- plot1 + ggplot2::scale_x_log10()
plot1 <- plot1 + ggplot2::scale_colour_manual(values = c(`FDR < 0.05` = "#eb6834",
                                                         `not significant` = "#9e9d97"),
                                              name = NULL)
plot1 <- plot1 + ggplot2::scale_shape_manual(values = c(null = 16, DE = 17),
                                             name = "True status")
plot1 <- plot1 + ggplot2::labs(x = "devel: estimated nuisance rate (log scale)",
                               y = "|Welch statistic|",
                               title = "A diverged rate (right-hand cluster) inflates the Welch statistic")
plot(plot1)
```

```{r diverged_table}
devel_df <- gene_df[gene_df$version == "devel", ]
diverged_df <- do.call(rbind, lapply(regime_levels, function(r){
  x <- devel_df[devel_df$regime == r & !devel_df$is_de, ]
  fp_vec <- x$fdr < 0.05
  data.frame(regime = r,
             frac_diverged = mean(x$nuisance > 1e4),
             fp_rate_diverged = mean(fp_vec[x$nuisance > 1e4]),
             fp_rate_other = mean(fp_vec[x$nuisance <= 1e4]))
}))
knitr::kable(diverged_df, digits = 3,
             col.names = c("Regime", "Fraction of null genes with rate > 1e4",
                           "FDR < 0.05 rate, diverged", "FDR < 0.05 rate, others"),
             caption = "devel only: false-discovery rate among null genes, split by whether the nuisance estimate diverged.")
```

## The t → z mapping (`NEWS`: "no longer produces an infinite statistic")

`master` computed the Gaussian statistic as `qnorm(pt(T, df))`, which returns
`Inf` once `pt()` rounds to 1. `devel` works on the log scale. **In these
simulations `master` never reached `Inf`**, because its capped nuisance rates
keep its Welch statistics below about 26. Applying the old mapping to
**`devel`'s** statistics shows where it would have failed:

```{r t_to_z_table}
tz_df <- do.call(rbind, lapply(c("master", "devel"), function(v){
  x <- gene_df[gene_df$version == v, ]
  z_old_vec <- stats::qnorm(stats::pt(x$teststat, df = x$df))
  data.frame(version = v,
             max_teststat = max(x$teststat),
             num_inf_old_mapping = sum(!is.finite(z_old_vec)),
             num_nonfinite_reported = sum(!is.finite(x$gaussian_teststat)))
}))
knitr::kable(tz_df, digits = 1,
             col.names = c("Statistics from", "Largest Welch statistic",
                           "Inf under master's mapping",
                           "Non-finite z actually reported"),
             caption = "All 60 data sets.")
```

So the log-scale mapping is a real fix, but the situation it guards against
arises chiefly because of the other change: the uncapped nuisance rate
produces the huge statistics.

## Which empirical-null estimator ran

`devel` records this; `master` does not (`method` is `NA`).

```{r method_table}
method_df <- run_df[run_df$version == "devel", ]
method_tab <- table(factor(method_df$regime, levels = regime_levels),
                    method_df$method)
knitr::kable(as.data.frame.matrix(method_tab),
             caption = "devel: estimator used, out of 10 replicates per regime.")
```

```{r null_param_plot, fig.width=9, fig.height=3.5}
plot_df <- run_df[run_df$version %in% c("master", "devel"), ]
plot_df$regime <- factor(plot_df$regime, levels = regime_levels)
plot_df$version <- factor(plot_df$version, levels = c("master", "devel"))
plot1 <- ggplot2::ggplot(plot_df, ggplot2::aes(x = version, y = null_sd))
plot1 <- plot1 + ggplot2::geom_point(ggplot2::aes(colour = version),
                                     position = ggplot2::position_jitter(width = 0.15,
                                                                         height = 0,
                                                                         seed = 1),
                                     size = 2)
plot1 <- plot1 + ggplot2::facet_wrap(~ regime, nrow = 1, scales = "free_y")
plot1 <- plot1 + ggplot2::scale_colour_manual(values = version_col_vec,
                                              guide = "none")
plot1 <- plot1 + ggplot2::labs(x = NULL, y = "Estimated null SD",
                               title = "Empirical-null SD per replicate")
plot(plot1)
```

`devel` falls back from `locfdr` to the truncated MLE in a minority of
replicates, most often in `null` and `strong_de`, and says so with a warning.
`master` may have fallen back just as often; it had no way to report it.

# Corner cases

Each row below perturbs one small cohort (10 individuals × 30 cells, 100 genes,
10 DE) and runs the full pipeline. "silent" means the pipeline finished
without warnings.

```{r corner_table}
case_vec <- unique(corner_df$case)
corner_wide_df <- do.call(rbind, lapply(case_vec, function(cs){
  m <- corner_df[corner_df$case == cs & corner_df$version == "master", ]
  d <- corner_df[corner_df$case == cs & corner_df$version == "devel", ]
  .describe <- function(x){
    txt <- x$outcome
    if(x$outcome == "error") txt <- paste0("**error** in ", x$error)
    if(x$outcome == "warning") txt <- paste0("warning: ", x$warnings)
    if(!is.na(x$num_sig)) txt <- paste0(txt, " — ", x$num_sig, " genes at FDR < 0.05")
    if(!is.na(x$metric_value)) txt <- paste0(txt, "; ", x$metric_name, " = ",
                                             signif(x$metric_value, 3))
    substr(txt, 1, 330)
  }
  data.frame(Case = paste0("`", cs, "`: ", d$description),
             master = .describe(m),
             devel = .describe(d))
}))
knitr::kable(corner_wide_df,
             caption = "Behavior of each version on the perturbed cohorts. Messages truncated.")
```

Reading the table:

* **Fixed, and clearly better.** `collinear_covariate` and `confounded_design`
  (Sex identical to case-control status) used to fail two stages later with
  `missing value where TRUE/FALSE needed` from `opt_esvd()`. Now they stop in
  `initialize_esvd()`, which names the redundant column. `all_zero_gene` used
  to finish with glmnet convergence warnings and fit the zero genes anyway;
  now it is refused by name and pointed at `eSVD_helper()`.
  `one_case_individual` used to die in `compute_pvalue()` on `NaN`s; now it
  stops in `compute_test_statistic()` with an explanation.
  `two_cell_individual` used to run silently; now it is refused (it can be
  overridden with `min_cells_per_individual = 0`).
* **`few_genes` is the most striking row.** With 30 genes, `master` declared
  **every gene significant**, silently. Its unconstrained fallback estimator
  gave a degenerate empirical null. `devel`'s `locfdr` converged here and
  called nothing significant, which is conservative but at least not
  nonsense. (Across the head-to-head regimes `devel` does fall back in some
  replicates, and it warns when it does.)
* **Unchanged.** `individual_in_both_arms` is refused by the same
  `stopifnot()` in both versions. The friendlier check lives in `eSVD()`,
  which these scripts do not call.
* **Not a practical difference in these data.** `determinism`: `master`'s
  two fits after different `set.seed()` calls differed by about 1e-11 in the
  Welch statistic, so its random SVD start rarely mattered here; `devel` is
  bit-identical. `sparse_input` agrees with dense input in both.
* **`sparse_na` shows a gap between `NEWS.md` and behavior.** `NEWS` says
  `initialize_esvd()` now zeroes `NA` counts in sparse matrices, but the
  pipeline never gets that far: `format_covariates()` computes
  `Log_UMI = log(rowSums(dat))`, which is `NA` for that cell, and
  `initialize_esvd()` then stops on the covariates. The error is clearer
  than `master`'s (`covariates` contains NA versus `missing value where
  TRUE/FALSE needed`), but it names the covariates rather than the count
  that caused them. The zeroing only helps a caller who builds the covariates
  some other way.
* **`extreme_de`** (five genes up by $e^4$ on nearly Poisson genes): both
  versions find the five genes. Neither overflowed at this size.

# New in devel: a standard error for the log2 fold change

`master` has no counterpart. Below, the interval `logFC ± 2 SE` is checked
against the true log2 fold change, split by whether the gene's nuisance rate
diverged.

```{r se_table}
se_df <- do.call(rbind, lapply(regime_levels, function(r){
  x <- devel_df[devel_df$regime == r, ]
  cover_vec <- abs(x$logFC - x$truth_log2fc) <= 2 * x$logFC_se
  diverged_vec <- x$nuisance > 1e4
  data.frame(regime = r,
             coverage_ordinary = mean(cover_vec[!diverged_vec]),
             coverage_diverged = ifelse(any(diverged_vec),
                                        mean(cover_vec[diverged_vec]), NA),
             median_se = stats::median(x$logFC_se))
}))
knitr::kable(se_df, digits = 3,
             col.names = c("Regime", "Coverage of ±2 SE, ordinary rate",
                           "Coverage of ±2 SE, diverged rate", "Median SE"),
             caption = "devel only, all genes and replicates.")
```

```{r se_plot, fig.width=8, fig.height=4.5}
plot_df <- devel_df[devel_df$regime == "baseline" & devel_df$rep == 1, ]
plot_df$diverged <- ifelse(plot_df$nuisance > 1e4, "rate diverged", "ordinary")
plot1 <- ggplot2::ggplot(plot_df, ggplot2::aes(x = truth_log2fc, y = logFC))
plot1 <- plot1 + ggplot2::geom_abline(slope = 1, intercept = 0,
                                      colour = "#c3c2b7", linewidth = 0.5)
plot1 <- plot1 + ggplot2::geom_errorbar(ggplot2::aes(ymin = logFC - 2 * logFC_se,
                                                     ymax = logFC + 2 * logFC_se,
                                                     colour = diverged),
                                        width = 0, alpha = 0.5)
plot1 <- plot1 + ggplot2::geom_point(ggplot2::aes(colour = diverged), size = 1.2)
plot1 <- plot1 + ggplot2::scale_colour_manual(values = c(ordinary = "#2a78d6",
                                                         `rate diverged` = "#eb6834"),
                                              name = NULL)
plot1 <- plot1 + ggplot2::labs(x = "True log2 fold change",
                               y = "devel: log2 fold change ± 2 SE",
                               title = "baseline, replicate 1")
plot(plot1)
```

For genes with an ordinary nuisance estimate, the SE covers the truth at or
above the nominal rate in five of the six regimes: it is conservative, as
`CLAUDE_kevin.md` notes, because the within-individual term dominates. The
exception is `strong_de` (coverage 0.84). There, the ±3 effects move each
case cell's total count, so the `Log_UMI` depth adjustment shifts every
gene's fold change. That is the bias recorded in `CLAUDE_kevin.md`, which an
SE cannot absorb. For genes whose rate diverged,
the interval collapses and coverage falls well below nominal. This is the
same failure as in section 6, seen from the SE's side.

# Runtime

```{r runtime_table}
runtime_df <- stats::aggregate(seconds ~ regime + version, data = run_df,
                               FUN = stats::median)
runtime_wide_df <- stats::reshape(runtime_df, idvar = "regime",
                                  timevar = "version", direction = "wide")
knitr::kable(runtime_wide_df, digits = 2,
             col.names = c("Regime", "devel (s)", "devel_swap (s)", "master (s)"),
             caption = "Median seconds per fit (600 cells × 300 genes).")
```

`devel` is about 35-55% slower on these small fits (about 0.6 s per fit).
The cost has not been profiled, so where it comes from is not known.

# Verdict

* **On the estimate researchers report first, the fold change, nothing
  changed.**
* **On the test, one change is dramatic, and it is not a corner-case
  change.** Uncapping `gamma_rate` gives nuisance rates that are more
  faithful in rank but biased upward, and occasionally divergent. That
  shrinks posterior variances and inflates Welch statistics. In ordinary
  regimes this costs a few false discoveries per 300 genes (from 0.2–1.8 up
  to 2.7–3.8 per data set). In low-overdispersion data it breaks
  calibration (type-I error 0.28 at p < 0.05). The ablation attributes all
  of it to this one change.
* **Everything else in `NEWS.md` behaves as a corner-case fix**, and in
  those corners `devel` is clearly better: clear, early errors instead of
  deep crashes, and no silent "everything is significant" on small gene
  sets.

**What this suggests for the CRAN release** (Kevin's decision; nothing here
has been changed): the `gamma_rate` change needs either a guard or an
upper bound before release. The bound would be a principled cap on
$\beta_j$, or a flag in `report_results()` for genes whose rate ran to the
edge of the search. Q10 and Q-LFC-1 in `CLAUDE_kevin.md` are the open
questions this data speaks to. The ablation machinery here (`devel_swap`)
can test any candidate fix in about three minutes.

# What we glossed over

* **Only simulated data, and mostly data from the eSVD model itself.** The
  one outside generator (`generate_null()`) shows the same pattern in
  milder form. Real cohorts have thousands of genes, which helps `locfdr`
  and may change the size of the tail effect.
* **300 genes, 20 individuals, 30 cells per individual.** Chosen to fit the
  30-minute budget. The false-discovery counts scale with the number of
  genes; the rates are what to compare.
* **`eSVD()` itself was not run**, because it needs a Seurat object. The
  scripts call its steps in the same order and with its defaults, except
  `k = 3` and `max_iter = 50`. `compute_test_per_gene()` (the
  `bool_diet = TRUE` path) was not compared.
* **The rescaling by `bool_stabilize_underdispersion`** divides every rate by
  their geometric mean when that mean is above 1. It is on in both versions,
  and diverged genes inflate that mean under `devel`. Its separate
  contribution was not isolated.

# References

* Lin, K. Z., Qiu, Y., and Roeder, K. (2024). eSVD-DE: cohort-wide
  differential expression in single-cell RNA-seq data using exponential-family
  embeddings. *BMC Bioinformatics* 25:113.
  [doi:10.1186/s12859-024-05724-7](https://doi.org/10.1186/s12859-024-05724-7)
* `NEWS.md` of this repository, sections 1.1.0 and 1.0.2.
