---
title: "eSVD-DE Overview"
author: "Kevin Z. Lin"
date: "2024-06-15"
output: 
  rmarkdown::html_vignette:
    df_print: "kable"
vignette: >
  %\VignetteIndexEntry{eSVD-DE Overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
collapse = TRUE,
comment = "#>"
)
```

# Purpose

The simulations here are meant to be portable, toy-examples of eSVD2, with the intent to demonstrate that an installation of eSVD2 was successful.

A successful installation of `eSVD2` is required for this. See the last section of this vignette for the session information used when it was rendered. The whole vignette should complete in less than 2 minutes.

# Overview

Broadly speaking, eSVD-DE (the name of the method, which is implemented in the `eSVD2` package) is a pipeline of 6 different function calls. The procedure starts with primarily 3 inputs, 

- `dat`, either `matrix` or `dgCMatrix`, with `n` rows (for `n` cells) and `p` columns (for `p` genes). We advise making sure each row/column of either matrix has non-zero variance prior to using this pipeline.
- `covariates`, a `matrix` with `n` rows with the same rownames as `dat` where the columns represent the different covariates Notably, this should contain only numerical columns (i.e., all categorical variables should have already been split into numerous indicator variables), and all the columns in `covariates` will (strictly speaking) be included in the eSVD matrix factorization model.
- `metadata_individual`, a `factor` vector with length `n` which denotes which cell originates from which individual.

```{r, message = FALSE}
library(eSVD2)
# Seurat is only used for the UMAP below; every eSVD2 step runs without it.
has_seurat <- requireNamespace("Seurat", quietly = TRUE)
# ggplot2 and ggrepel are only used for the diagnostic plots at the end.
has_ggplot <- requireNamespace("ggplot2", quietly = TRUE) &&
  requireNamespace("ggrepel", quietly = TRUE)
```


# Simulating data

We create a simple dataset that has no differentially expressed genes via the `eSVD2::generate_null()` function.

```{r}
set.seed(10)
res <- eSVD2::generate_null()
covariates <- res$covariates
metadata_individual <- res$metadata_individual
obs_mat <- res$obs_mat
```

If you have Seurat also installed, you can run the following lines as well (the chunk is skipped otherwise),

```{r, eval = has_seurat, warning = FALSE, message = FALSE, fig.width = 6, fig.height = 6}
set.seed(10)
meta_data <- data.frame(covariates[, c("CC", "Sex", "Age")])
meta_data$Individual <- metadata_individual
seurat_obj <- Seurat::CreateSeuratObject(counts = Matrix::t(obs_mat), meta.data = meta_data)
Seurat::VariableFeatures(seurat_obj) <- rownames(seurat_obj)
seurat_obj <- Seurat::NormalizeData(seurat_obj)
seurat_obj <- Seurat::ScaleData(seurat_obj)
seurat_obj <- Seurat::RunPCA(seurat_obj, verbose = FALSE)
seurat_obj <- Seurat::RunUMAP(seurat_obj, dims = 1:5)

Seurat::DimPlot(seurat_obj, reduction = "umap", group.by = "CC")
```

The separation between case and control cells is purely determined by the individual covariates of Age and Sex, not the case-control status.

# Using eSVD-DE

* **Step 1 (Initializing the eSVD object):** First, we initialize the object of class `eSVD`, which we will call `eSVD_obj`. This is done using the `eSVD2::initialize_esvd()` function. All the following operations will use
`eSVD_obj` for all calculations through the usage of this method. This is where we pass the inputs `dat`, `covariates`, and `metadata_individual`. Notice that after `eSVD2::initialize_esvd()` (and later, after `eSVD2::opt_esvd()`) we always call `eSVD2::reparameterization_esvd_covariates()` to ensure that the fit remains identifiable.

Empirically, we have found it useful to set `omitted_variables` as the `Log_UMI` and the batch variables. (Here, there are no batch variables.) This means that when eSVD2 reparameterizes the variables, it does not fiddle with the coefficient for `Log_UMI`. We found this beneficial since the coefficient for specifically the covariate `Log_UMI` bears a special meaning (as it impacts the normalization of the sequencing depth most directly), and we do not wish to "mix" covariates correlated with `Log_UMI` too early. (Overall, this is not a strict recommendation, however, we have found this practice to be beneficial in almost all instances we have used eSVD2, so we ourselves have not tried deviating from this convention.)

This typically looks like the following.
```{r}
set.seed(10)
eSVD_obj <- eSVD2::initialize_esvd(
  dat = obs_mat,
  covariates = covariates,
  metadata_individual = metadata_individual,
  case_control_variable = "CC",
  bool_intercept = TRUE,
  k = 5,
  lambda = 0.1,
  verbose = 0
)
eSVD_obj <- eSVD2::reparameterization_esvd_covariates(
  input_obj = eSVD_obj,
  fit_name = "fit_Init",
  omitted_variables = "Log_UMI"
)

names(eSVD_obj)
```

* **Step 2 (Fitting the eSVD object):** 
We now fit the eSVD. Typically, we recommend fitting twice
(i.e., calling `eSVD2::opt_esvd()` twice) due to the non-convex nature of the objective function. In the first call of `eSVD2::opt_esvd()`, hold all the coefficients to all the covariates aside from the case-control covariate (here, denoted as `"CC"` so we remove as much of the case-control covariate effect as possible. Then, in the second call of `eSVD2::opt_esvd()`, we allow all the coefficients to be optimized.

Note that in the last fit of `eSVD2::opt_esvd()`, we do not set any `omitted_variables`. This is our typical recommendation -- it is good to allow all coefficients to be reparameterized right before the optimization is all done.

```{r}
set.seed(10)
eSVD_obj <- eSVD2::opt_esvd(
  input_obj = eSVD_obj,
  l2pen = 0.1,
  max_iter = 100,
  offset_variables = setdiff(colnames(eSVD_obj$covariates), "CC"),
  tol = 1e-6,
  verbose = 0,
  fit_name = "fit_First",
  fit_previous = "fit_Init"
)
eSVD_obj <- eSVD2::reparameterization_esvd_covariates(
  input_obj = eSVD_obj,
  fit_name = "fit_First",
  omitted_variables = "Log_UMI"
)

set.seed(10)
eSVD_obj <- eSVD2::opt_esvd(
  input_obj = eSVD_obj,
  l2pen = 0.1,
  max_iter = 100,
  offset_variables = NULL,
  tol = 1e-6,
  verbose = 0,
  fit_name = "fit_Second",
  fit_previous = "fit_First"
)

eSVD_obj <- eSVD2::reparameterization_esvd_covariates(
  input_obj = eSVD_obj,
  fit_name = "fit_Second",
  omitted_variables = NULL
)

names(eSVD_obj)
```

* **Step 3 (Estimating the nuisance (i.e., over-dispersion) parameters):** 
After fitting the eSVD, we need to estimate the nuisance parameters. Since we are modeling the data via a negative binomial, this nuisance parameter is the over-dispersion parameter.
This is done via the `eSVD2::estimate_nuisance()` function.

```{r}
set.seed(10)
eSVD_obj <- eSVD2::estimate_nuisance(
  input_obj = eSVD_obj,
  bool_covariates_as_library = TRUE,
  bool_library_includes_interept = TRUE,
  bool_use_log = FALSE,
  cap_multiplier = 10,
  min_val = 1e-4,
  verbose = 0
)

names(eSVD_obj)
```

The value stored in `eSVD_obj$fit_Second$nuisance_vec` is the Gamma *rate*, the reciprocal of the over-dispersion, so a larger value means less over-dispersion. **No gene's rate is allowed above `cap_multiplier` times the median of its library size.** A gene whose counts are no more variable around the fit than a Poisson has no finite maximum-likelihood rate, and without the cap its test statistic is inflated. The function records what it did with each gene, and how many genes the cap acted on.

```{r nuisance_status}
table(eSVD_obj$fit_Second$nuisance_status)
eSVD_obj$param$nuisance_num_capped
```


* **Step 4 (Computing the posterior):** 
With the over-disperison parameter estimated, we are now ready to compute the posterior mean and variance of each cell's gene expression value. This posterior distribution is to balance out the fitted eSVD denoised value against the covariate-adjusted library size. This step ensures that even though we pooled all the genes to denoise the single-cell expression matrix, we do not later inflate the Type-1 error. This is done via the `eSVD2::compute_posterior()` function.

```{r}
set.seed(10)
eSVD_obj <- eSVD2::compute_posterior(
  input_obj = eSVD_obj,
  bool_adjust_covariates = FALSE,
  alpha_max = 100,
  bool_covariates_as_library = TRUE,
  bool_stabilize_underdispersion = TRUE,
  library_min = 1,
  pseudocount = 0
)

names(eSVD_obj)
```

* **Step 5 (Computing the test statistics):** 
Now we can compute the test statistics using the `eSVD2::compute_test_statistic()` function.

```{r}
set.seed(10)
eSVD_obj <- eSVD2::compute_test_statistic(input_obj = eSVD_obj, verbose = 0)

names(eSVD_obj)
```

* **Step 6 (Computing the p-values):** 
Lastly, we can compute the p-values via the `eSVD2::compute_pvalue()` function.

```{r}
set.seed(10)
eSVD_obj <- eSVD2::compute_pvalue(input_obj = eSVD_obj)

names(eSVD_obj)
```

To visualize the p-values, we can plot the theoretical quantiles against the observed quantiles. We see that despite the UMAP above showing strong differences between the case and control cells, the p-values are all uniformly distributed. This shows that we do not inflate the Type-1 error.

```{r, fig.width = 6, fig.height = 6}
pvalue_vec <- 10 ^ (-eSVD_obj$pvalue_list$log10pvalue)
graphics::plot(
  x = sort(pvalue_vec),
  y = seq(0, 1, length.out = length(pvalue_vec)),
  asp = TRUE,
  pch = 16,
  xlab = "Observed quantile",
  ylab = "Theoretical quantile"
)
graphics::lines(
  x = c(0, 1),
  y = c(0, 1),
  col = "red",
  lty = 2,
  lwd = 2
)
```

* **Step 7 (Reporting the results):**
The function `eSVD2::report_results()` collects the results into a data frame with one row per gene. The column `logFC` is the log2 fold change between the case and the control individuals, and `logFC_se` is its standard error, also on the log2 scale. Both are computed during Step 5 and are stored in `eSVD_obj$log2fc_vec` and `eSVD_obj$log2fc_se_vec`.

```{r}
result_df <- eSVD2::report_results(eSVD_obj)
utils::head(result_df[order(result_df$log10pvalue, decreasing = TRUE), ])
```

The standard error is divided by the number of individuals in each arm and not by the number of cells, so it is on the same footing as the standard error from a pseudobulk analysis. Multiply `logFC` and `logFC_se` by `log(2)` to move both to the natural-log scale. Note that `logFC / logFC_se` is not the statistic behind the p-value, which is computed on the linear scale and calibrated with an empirical null. See `?eSVD2::compute_log_fold_change` for what the standard error does and does not account for.

# Diagnostics of the over-dispersion

The over-dispersion decides how much each cell's posterior follows the fit, so it is worth a look before the results are used. The two functions below draw what Step 3 estimated. They need the packages `ggplot2` and `ggrepel`.

**The rate of each gene, against three summaries of the gene.** `eSVD2::plot_nuisance()` draws the rate divided by the gene's median library size, which puts every gene in the same units. The dashed line is the cap, and the genes it acted on sit on it. Most genes should be below the line. Where genes are at the line, they should be the lowly expressed and sparse ones, and they should not make up the most significant genes.

```{r plot_nuisance_expression, eval = has_ggplot, fig.width = 7, fig.height = 5}
plot1 <- eSVD2::plot_nuisance(input_obj = eSVD_obj,
                              x_axis = "mean_expression")
plot(plot1)
```

```{r plot_nuisance_pvalue, eval = has_ggplot, fig.width = 7, fig.height = 5}
plot1 <- eSVD2::plot_nuisance(input_obj = eSVD_obj,
                              x_axis = "log10pvalue")
plot(plot1)
```

```{r plot_nuisance_sparsity, eval = has_ggplot, fig.width = 7, fig.height = 5}
plot1 <- eSVD2::plot_nuisance(input_obj = eSVD_obj,
                              x_axis = "sparsity")
plot(plot1)
```

**The fitted count against the observed count.** In `eSVD2::plot_fitted_vs_observed()` each point is one cell and one gene, and a point is red when its count is more than 3 standard deviations from its fit, the standard deviation being the one the model implies for that gene's over-dispersion. The points should follow the dashed line, and only a small share should be red. With `bool_draw_bars = TRUE` each point carries the interval of 3 standard deviations around its fit: bars that are too short to reach the line mean the over-dispersion is underestimated, and bars that reach far past it for nearly every point mean it is overestimated.

```{r plot_fitted_all, eval = has_ggplot, fig.width = 7, fig.height = 5}
plot1 <- eSVD2::plot_fitted_vs_observed(input_obj = eSVD_obj,
                                        max_points = 5000)
plot(plot1)
```

The argument `genes` draws every cell of the genes named, one panel per gene.

```{r plot_fitted_genes, eval = has_ggplot, fig.width = 7, fig.height = 4}
plot1 <- eSVD2::plot_fitted_vs_observed(input_obj = eSVD_obj,
                                        genes = colnames(obs_mat)[1:2],
                                        bool_draw_bars = TRUE)
plot(plot1)
```

**Redoing the test at another cap.** The factorization does not depend on the cap, so the test can be redone at another `cap_multiplier` without fitting again. `cap_multiplier = 1` is close to the behavior of the version of `eSVD2` that accompanied the paper, and `Inf` removes the cap.

```{r recompute_pvalue, warning = FALSE}
eSVD_obj_cap1 <- eSVD2::recompute_pvalue(input_obj = eSVD_obj,
                                         cap_multiplier = 1)
eSVD_obj_cap1$param$nuisance_num_capped
stats::cor(eSVD_obj$teststat_vec, eSVD_obj_cap1$teststat_vec)
```

> **When the object has no counts.** `eSVD2::eSVD()` and `eSVD2::eSVD_helper()` return an object without the count matrix by default (`bool_diet = TRUE`). For such an object, pass the Seurat object the analysis was run on as `seurat_obj` to `eSVD2::recompute_pvalue()` and to `eSVD2::plot_fitted_vs_observed()`. `eSVD2::plot_nuisance()` needs no counts.

# Setup

The session information when this vignette was rendered.

```{r}
sessionInfo()
```
